---
title: 'Reproducible Research: Peer Assessment 1'
output: 
  html_document:
    keep_md: yes
---

```{r, include=FALSE}
library(dplyr)
library(ggplot2)
library(lattice)
library(crayon)
```

## **Loading and preprocessing the data**

Unzip and load data
```{r}
unzip(zipfile = 'activity.zip')
activ_data <- read.csv('activity.csv')
```

Process and transform data
```{r}
#format date var
activ_data$date <- as.POSIXct(activ_data$date, format = '%Y-%m-%d')
```

Create new data frame grouped by total steps per day
```{r}
total_steps <- group_by(activ_data[complete.cases(activ_data),], date) %>% summarise(total_steps = sum(steps, na.rm = TRUE))
```

Create histogram of the total number of steps taken each day
```{r, echo=FALSE}
ggplot(total_steps, aes(x = total_steps)) +
  geom_histogram(color = 4, fill = "white", bins = 30) + 
  xlab('Total steps per day') +
  ylab('Frequency (count)') 
```

## **What is mean total number of steps taken per day?**
```{r}
cat(bold('The mean total steps per day is:', mean(total_steps$total_steps)))

cat(bold('The median total steps per day is:', median(total_steps$total_steps)))
```

## **What is the average daily activity pattern?**
Create new data set
```{r}
ts_data <- group_by(activ_data, interval) %>% 
   summarise(mean_steps = mean(steps, na.rm = T))
```

Plot the data
```{r, echo=FALSE}
plot(ts_data$interval, ts_data$mean_steps, type="l", col = "blue", xlab="Interval", ylab="Average Number of Steps",main="Activity by Interval")
```

Calculate a 5-minute interval that, on average, contains the maximum number of steps
```{r}
max_interval <- ts_data[which.max(ts_data$mean_steps),1]
max_interval <- unlist(max_interval);
cat("5-min interval with the maximum number of steps:", max_interval)
```

## **Imputing missing values**

Calculate and report the total number of missing values in the dataset 
```{r}
cat('Number of NAs in steps:', sum(is.na(activ_data$steps)))

cat('Number of NAs in date:', sum(is.na(activ_data$date)))

cat('Number of NA in inteval:', sum(is.na(activ_data$interval)))
```

The strategy used to account for missing data was to impute the mean for any NA.

```{r}
activ_data$steps[is.na(activ_data$steps)] <- mean(activ_data$steps, na.rm = T)  
cat('Number of NA in steps post imputation:', sum(is.na(activ_data$steps)))
```

Create histogram of the total number of steps taken each day after missing values are imputed

```{r}
new_total_steps <- group_by(activ_data[complete.cases(activ_data),], date) %>% summarise(total_steps = sum(steps, na.rm = TRUE))
```

Plot the data
```{r, echo=FALSE}
ggplot(new_total_steps, aes(x = total_steps)) +
  geom_histogram(color = 4, fill = "white", bins = 30) + 
  xlab('Total steps per day') +
  ylab('Frequency (count)') 
```

## **Are there differences in activity patterns between weekdays and weekends?**

Create new variable with two levels: weekday and weekend
```{r}
activ_data$date <- as.Date(activ_data$date)
activ_data$weekday <- weekdays(activ_data$date)
activ_data$weekday = ifelse(activ_data$weekday %in% c('Sunday', 'Saturday'), 'Weekend', 'Weekday')
```

Create new data frame grouped by total steps per day
```{r}
new_ts_data <- activ_data %>% group_by(weekday, interval) %>% 
   summarise(mean_steps = mean(steps))
```

Plot data
```{r, echo=FALSE}
xyplot(new_ts_data$mean_steps ~ new_ts_data$interval|new_ts_data$weekday, groups = new_ts_data$weekday, main="Activity by Day of the Week",xlab="5-Minute Interval", ylab="Average Number of Steps",layout=c(1,2), type="l", col=c("red","blue"))
```

```{r}
wk_grp_mean <- new_ts_data %>% group_by(new_ts_data$weekday) %>% summarise(across(mean_steps, mean, na.rm = TRUE))

cat(bold('Activity Patterns for Weekends are higher than Weekdays.'))
cat(bold('The mean total steps for weekdays is:', unlist(wk_grp_mean[1,2])))
cat(bold('The mean total steps for weekend is:', unlist(wk_grp_mean[2,2])))

```
